TBD

### 2021-02-26 outputs/paper/paper.Rmd
Added written components (abstract, introduction, and discussion)

---
title: "TBD"
subtitle: "TBD"
author: "Najma Osman, Rachit Srivastava, Morgaine Westin, and Bongju Yoo"
thanks: "Code and data are available at: https://github.com/bonjwow/covid-shut-down"
date: "`r format(Sys.time(), '%d %B %Y')`"
abstract: "First sentence. Second sentence. Third sentence. Fourth sentence."
output:
  bookdown::pdf_document2:
    fig_caption: yes
geometry: margin=0.75in
header-includes:
- \usepackage{float} #use the 'float' package
- \floatplacement{figure}{H} #make every figure with caption = h
- \usepackage{longtable}
toc: FALSE
bibliography: references.bib
---

@@

@@

```{r Q4, echo = FALSE, message = FALSE, fig.cap = "Number of years restaurants have been in operation"}
### How many years has your restaurant been in operation?
### 1-4 years
### 5-7 years
### 8-10 years
### 10 or more years
q4 <- ggplot(data = all_simulated_dataset, mapping = aes (x = Q4, fill = type)) +
  geom_bar() + coord_flip() +
  scale_x_discrete(limit = c ("3 years or less", "5-7 years", "8-10 years", "10 or more years")) +
  labs(x = "Years of Operation",
       y = "# of Restaurants",
       fill = "Condition",
       title = str_wrap("Number of Years in Operation", 30)) +
  theme_fivethirtyeight() +
  scale_fill_tableau(labels = c("Control", "Intervention"), guide = guide_legend(reverse = TRUE)) + 
  theme(title=element_text(size=9))
```

```{r samplechar, fig.width=9, fig.height=7, echo = FALSE, message = FALSE, fig.cap="Sample Characteristics of Restaurants by Condition"}
grid.arrange(q1, q2, q3, q4)
```

@@

@@

>>>>

---
title: "TBD"
subtitle: "TBD"
author: "Najma Osman, Rachit Srivastava, Morgaine Westin, and Bongju Yoo"
thanks: "Code and data are available at: https://github.com/bonjwow/covid-shut-down"
date: "`r format(Sys.time(), '%d %B %Y')`"
abstract: "In light of recent COVID-19 restrictions and its impact on businesses, a randomized control trial was conducted by Petit Poll on behalf of the Ontario Government to test the effects of shutting down indoor dining services on restaurants in Toronto. Restaurants who received the intervention and only offered take-out or delivery services experienced a decline in their quarterly revenues compared to the previous quarter and had to make adjustments to the number of employees, menu prices, and operational schedules of their restaurants. Conversely, restaurants in the control group whose indoor dining services remained open during the study period experienced no significant changes in their revenue, employee count, menu prices, or operational schedule compared to the previous corner. While food delivery and take-out services are becoming more popular, these results indicate the value of indoor dining experiences for both restaurants and customers, which the Ontario Government can take into consideration when determining how they can best support Toronto’s growing restaurant industry. "
output:
  bookdown::pdf_document2:
    fig_caption: yes
geometry: margin=0.75in
header-includes:
- \usepackage{float} #use the 'float' package
- \floatplacement{figure}{H} #make every figure with caption = h
- \usepackage{longtable}
toc: FALSE
bibliography: references.bib
---

@@

Toronto is a cluster of diverse communities with many different aspects that make it a unique place to live. One of these aspects include restaurant businesses that consist of a wide variety of cuisines all across the city. The Ontario government department has asked us, Petit Poll, to design an experiment to test the implications of COVID-19 shutdowns and its effect on restaurant businesses. During Toronto’s tightest lockdown measures, many restaurants were forced to establish ghost kitchens in which their facilities were designed to focus exclusively on take-out and delivery orders [@kelly2019]. While some restaurants had already been using the ghost kitchen model before the pandemic, the circumstances of the pandemic undoubtedly lead to a rise in this practice [@Doradea2020]. Focusing on take-out and delivery services reduces the costs necessary to run a fully operational brick and mortar restaurant and removes the need for many customer-facing staff members [@kelly2019]. However, there are potential drawbacks that come with this shift as customers are losing the benefits and experiences of indoor dining  while restaurants are losing an important source of revenue. As pandemic restrictions loosen and restaurants are able to open up their indoor dining services again, it is possible that many restaurants may make the decision to continue to focus exclusively on take-out and delivery. Given the unprecedented nature of this situation, we do not yet fully understand the extent to which this shift towards take-out and delivery only services will impact restaurants and customer behaviours in the long term. This study aims to measure the effects on restaurant performance which may be associated with shutting down indoor dining, and the changes that restaurants make to deal with this shift.

As data scientists for Petit Poll, we narrowed down our test to restaurants in the city of Toronto as it condenses our test results into one precise area within Ontario. By using a business licensing dataset obtained from the Toronto Open Data Portal [@todp2021], we were able to obtain information on all restaurants in the city. Since eating establishments have been impacted by COVID-19 in the past year, to which there have been irregular restrictions on the businesses, we believed this characteristic was best suited for this study. In Toronto, pandemic restrictions likely had an impact on customers’ dining behaviours, as they were forced to order restaurant food for take-out and delivery only. Now that restaurants are able to open up their full services again, will their dining habits shift as they have become accustomed to the convenience of food delivery? Exploring how shutting down indoor dining services again after they have just reopened may affect performance such as revenue can also provide insights on Torontonians’ dining habits and preferences in these regards.

This study was conducted for three months, from October 1st to December 31st, 2020 where we studied the comparison between a control group and an intervention group. The intervention group had to partake in the study by resuming their business as a take-out only business and a control group had to remain business as usual. This was done intentionally to maintain the integrity of the study by studying the differences between restaurants that closed their indoor dining services and the ones that did not. Survey data was collected via Google Forms to measure the effects of the intervention on restaurant’s employee numbers and revenue, and if changes were made in menu prices and operation days and hours during the intervention period. Survey data were processed and analyzed in R [@citeR], with a complete list of packages and their usages listed in the Appendix. Based on survey responses, it was revealed that most restaurants in the intervention group who shut down their indoor dining services experienced a decrease in their total sales revenue compared to the previous quarter, with many reporting an overall decline of 50% or more. During the study period, restaurants in the intervention group made several changes to their restaurant practices to account for these changes such as reducing the number of employees, adjusting their menu prices, and reducing their hours and days of operation. Meanwhile, the restaurants in the control group who continued operating at full capacity remained relatively stable in terms of their revenue, and did not make any significant changes to their employee counts, menu prices, or operational periods. This suggests that shutdowns of indoor dining practices due to COVID-19 have large implications on restaurant performance, where the revenue, hours and days of operation, and menu prices all get impacted to adjust to deviations from the restaurant’s original business model.

@@

```{r Q4, echo = FALSE, message = FALSE, fig.cap = "Number of years restaurants have been in operation"}
### How many years has your restaurant been in operation?
### 1-4 years
### 5-7 years
### 8-10 years
### 10 or more years
q4 <- ggplot(data = all_simulated_dataset, mapping = aes (x = Q4, fill = type)) +
  geom_bar() + coord_flip() +
  scale_x_discrete(limit = c ("3 years or less", "5-7 years", "8-10 years", "10 or more years")) +
  labs(x = "Years of Operation",
       y = "# of Restaurants",
       fill = "Condition",
       title = str_wrap("Number of Years in Operation", 30)) +
  theme_fivethirtyeight() +
  scale_fill_tableau(labels = c("Control", "Intervention"), guide = guide_legend(reverse = TRUE)) + 
  theme(title=element_text(size=9))
```
```{r samplechar, fig.width=9, fig.height=7, echo = FALSE, message = FALSE, fig.cap="Sample Characteristics of Restaurants by Condition"}
#combining plots for q1-4
grid.arrange(q1, q2, q3, q4)
```

@@

This study implemented a randomized control trial and used survey methodology to assess the effects of shutting down indoor dining services for restaurants in Toronto compared to a valid control group. Stratified random sampling and random assignment were used to select restaurants in Toronto to receive an intervention in which they were asked to rely only on take-out and delivery services for a three month period. Based on the survey results, we identified a few key insights that will help the Ontario government understand the implications of how restaurants were affected when they were forced to shut down their indoor dining services, how they tried to adapt to these changes, and what these results have to say about the dining habits of Torontonians. From the results, we found that the intervention had significant effects on restaurants compared to the control group. Notably, restaurants who received the intervention experienced a significant decrease in revenue compared to when they were operating with full services, while revenue remained stable for restaurants in the control condition. The intervention was also associated with a decrease in employees, adjustments in restaurants’ menu prices, and reduced hours and days of operation.

Proponents of the ghost kitchen model believe take-out and delivery focused businesses are here to stay as demands for food delivery are on the rise. [@Sundar2019]. Some may argue that the growth in popularity of delivery services, even before the pandemic, reflect changes in the population’s dining habits. The fast-paced society we live in may place a lot more value on the ease and convenience of food delivery than on dining inside a restaurant, which raises the question as to whether ghost kitchens are able to replace indoor dining services entirely. With that being said, while Torontonians enjoy the convenience of take-out and delivery, there is still value in keeping indoor dining experiences. The decreases in revenue observed in the intervention group suggest a significant portion of the restaurants previous source of revenue came from indoor dining. These restaurants had to adjust aspects of their business practices to cater to the intervention. By shutting down indoor dining and only allowing for deliveries and takeout orders, the restaurants had to make significant changes to their employee counts, menu price, and opening times. Despite these alterations, restaurants in the treatment group still had a decrease in their revenue compared to the previous quarter, while revenue in the control group remained stable. Based on this result, we can make an assumption that the dining habits of Torontontonians did not change even though the restaurants switched to a business model which highlighted the convenience of take-out and delivery. In fact, Torontonian’s require a mix of delivery services, takeout orders, and dining-in experiences at restaurants. Due to this reason, restaurants in the treatment group noticed a decline in their revenue during the time of the intervention. People want to dine-in at restaurants sometimes, so if they desire to dine-in and restaurants do not have that option then they will go to a restaurant which can offer those services to them.

Over the course of the pandemic, there were increased calls for Torontonians to support local restaurants and businesses [@Yelich2020]. However, mainstream delivery services such as UberEats and Doordash charge restaurants high commission fees up to 30%, meaning a significant portion of money made from orders goes to these companies and not the restaurants themselves [@Shea2021]. While commission fees for food delivery services in Ontario were capped at 15% in December 2020, during the last month of the intervention, some restaurants have opted to use their own delivery services to avoid the fees altogether. While our survey results indicated a small portion of restaurants used an in house delivery service, the majority of respondents still received the most orders from UberEats or DoorDash. An interview with the creator of Not Uber Eats, a commission free alternative to mainstream delivery services, reported that most customers in Toronto would prefer to see all their money go directly to the restaurant they are ordering from [@Shea2021]. If this is true, then it is possible that when customers wanted to eat at a restaurant but they only had the option to order the food through UberEats, they might have made the decision to dine indoors at a restaurant where they had a better idea of where all their money was going.

A portion of the revenue lost for restaurants in the intervention condition could also be attributed to additional sales commonly made as a result of the indoor dining experience. For instance, customers are more likely to be upsold or influenced by waitstaff to make other purchases such as appetizers, drinks, or desserts when dining inside a restaurant versus when they are ordering for delivery. Large social gatherings or celebrations which took place inside restaurants may have also resulted in more menu items being sold, and a greater likelihood of being upsold. In addition to this, an interesting finding was that shutting down indoor dining services caused many restaurants to adjust their menu prices in some way. While it would be fair to assume restaurants would raise their prices to account for revenue loss from indoor dining, most restaurants adjusted their menu prices by lowering them. Reduced menu prices could attract more customers, and entice them to add on more menu items to their orders. Price reductions may also reflect the decreased value customers are getting when they order take-out or delivery. Aside from the costs of food and preparation, menu prices are often designed to reflect the customer’s perceived value of a dish, and should reflect what consumers think is an appropriate price for that specific item [@shoemaker2005increase]. For customers, a great deal of the value of eating a restaurant meal comes from the indoor dining experience itself, such as the restaurant’s ambience, customer service, and social experiences [@Mohammed2020]. While price alterations reported in this study were likely only temporary changes, if more restaurants make the decision to permanently focus on only providing take-out and delivery, they should consider how customers’ perception of value for dining experiences may influence how much they are willing to pay for a meal.

Another thing to note for the Ontario government department is that the survey results implies that Toronto’s restaurant market is super saturated in fast-food services and not fine dining. This can help the Ontario government department invest more into fine-dining restaurants to bring awareness to these types of cuisines to Torontonians. If the residents want more options of cuisines that promote the notion of dining-in rather than taking-out food, then those options can be readily available for people if they desire. It can also be the opposite, where the department can use this result to invest into more fast-food services to bring more restaurants that are quick and convenient for people in Toronto. Based on the type of restaurant, the hours of operation and menu items can change accordingly to cater to the task of dining experience that people desire.

@@

There are several limitations associated with our study that must be addressed. The first involves the sampling methodology used and concerns about cross-contamination between restaurants in the intervention and control group. As discussed earlier, we relied heavily on the process of random sampling and the large pool of restaurants we were sampling from to circumvent the likelihood that restaurants in the intervention and control group would be in close proximity to one another and therefore impact each other's performance.  Since restaurants in each group were not separated by region, and were instead equally assigned within regions, it is possible that cross-contamination between groups still occurred, especially in smaller boroughs with less restaurants that are all within close proximity to each other. After the sample was selected and randomly assigned, we manually noted any restaurants in different groups which were in extreme close proximity to one another (i.e. directly next door or directly across the street), and fortunately found none. However, since it is difficult to estimate customer behaviours and the way they decide exactly which restaurants to eat or not eat at, there is still a possibility that some restaurant performances in the control group may have been directly impacted by the closures of restaurants in the intervention groups. Although given the number of restaurants available to eat at in Toronto, even in smaller areas, this is unlikely to have a large impact on the results.

Another limitation involves biases in our sample frame and response biases among restaurants who choose to participate in the experiment versus those who don’t. The sample of restaurants was drawn from a list of business licenses in the city of Toronto [@license2021], and restaurants whose licenses had been cancelled were filtered out from the sampling process. Unfortunately, a number of restaurants had to permanently close down due to COVID-19, a large portion of which were smaller scale businesses [@Xing2020]. A permanent shut-down of a restaurant would result in their business license being cancelled and updated on the dataset we used to collect our sample. This means that the frame from which our sample was selected was limited to the remaining restaurants who were able to “survive” the pandemic, majority of which were likely medium or large scale businesses. While our intentions for this study were to examine effects on a wide range of restaurant types and sizes, smaller restaurants may not have been represented in the experiment due to bias and limitations in the sampling frame.

In addition to this, response bias and non-response from restaurants who declined to participate in the study may also limit the generalizability of our results. Of the 200 restaurants randomly assigned to the intervention condition, 50 restaurants declined to participate. Unfortunately, there could be systematic differences between the restaurants who chose to participate compared to the restaurants that declined, meaning generalizability of our study results are limited to the characteristics of the former. There are a number of reasons why a restaurant owner would decline to participate in the study, the most likely one being the financial risks associated with shutting down their indoor dining services. It is understandable that asking smaller restaurants or restaurants who heavily rely on revenue from indoor dining to switch to a take-out only service for three months might be unreasonable and not worth the financial risks. This is why we tried to be upfront about any risks and potential losses which may occur, assured restaurants their participation was completely voluntary, and provided restaurants with the opportunity to opt-out at any point during the experiment. However, this could suggest that the restaurants who consented to participate in the intervention did so knowing that their restaurants already had infrastructure or services in place to succeed in a take-out only environment, or were confident in their ability to recuperate any financial losses experienced during the intervention. Again, this means there could be non-random differences in restaurant characteristics underlying the decision to participate in the study. A further analysis comparing the demographic characteristics of restaurants who chose to participate in the study with the characteristics of those who declined to participate might reveal areas where our sample might be biased. While our sampling methodology ensured equal representation of different restaurants among the intervention and control groups, it is unlikely that these characteristics are as evenly distributed between those who did and did not participate in the study. For instance, if all of the restaurants who declined to receive the intervention were small fine dining restaurants, that raises concerns that our study results were not as generalizable to all restaurant types as we had intended.

Finally, there are some limitations to the survey methodology we employed. The survey was designed to be short in length to avoid survey fatigue in respondents and mainly consisted of multiple choice questions to simplify analysis. While these methodology choices were deliberate on our end, they do come with several trade-offs. First, the short length of the survey meant we only focused on measuring a select few impacts that the intervention would have on restaurants, namely on employee numbers, revenue, menu prices, and hours and days of operation. However, it is possible that the intervention resulted in a number of other unforeseen effects on restaurants that the survey did not adequately capture. Our decisions to focus on measuring certain effects and not others, and the use of categorical forced-choice questions in the survey design limited our understanding of the full range of impacts that removing indoor dining services had on restaurants. Despite this, we were still able to gain valuable insights based on the effects we did measure, which will hopefully serve as a basis for future research on dining habits and restaurant business models.

Based on the limitations identified above, we have several suggestions for how our experiment can be extended and improved upon in future studies. Implementing surveys with open-ended questions or conducting interviews with restaurant owners will give us a better understanding of the breadth of effects that switching to a take-out and delivery only business model has on restaurants and identify any important effects we might have missed during the present study. Additionally, while the purpose of this experiment was to examine effects across all types of restaurants in Toronto, it might be beneficial to focus the intervention on certain types of restaurants perhaps based on size or segmentation in the future to see how the effects of the intervention may differ. For example, since only a small portion of small scale restaurants participated in the study, an intervention designed specifically to test the effects of implementing take-out or delivery only services in smaller businesses could be an area for future research. Of course, this would require sufficient funding and resources to be put in place to ensure an appropriate number of small restaurants would participate in the intervention without putting their businesses in extreme financial risk. Another potential avenue of exploration would be a more direct comparison of how different restaurant segmentation types impact the extent to which the intervention has an effect on restaurant performance. For instance, fine dining restaurants who rely more on customers who come in for indoor dining may have a harder time adapting to a ghost kitchen model than restaurants who were already geared towards quick and convenient service. A further examination of how the interaction between restaurant size and segmentation (i.e. small scale fine dining restaurants) impacts the intervention outcomes may also be useful. Understanding how different types of restaurants perform and adapt to take-out based business models will not only provide insights into the behaviours of restaurateurs and consumers, but will help identify areas where governments can provide additional support in situations where restaurants may once again be forced to shut down their indoor dining services. Finally, we made several claims about Torontonians’ dining habits and the value of indoor dining in this report. A more direct study of customers’ preferences for indoor dining versus delivery and take-out, and how they place value on indoor dining experiences, will expand on our findings and provide more insights into the future of dining practices in Toronto.
